---
title: Fitting a finite mixture of piecewise mixed-effect models with unknown random knots
subtitle:
author:
  name: Cengiz Zopluoglu
  affiliation: University of Oregon
date: 09/5/2024
output: 
  html_document:
    keep_md: false
    theme: journal
    highlight: haddock
    code_folding: show
    toc: yes
    toc_depth: 4
    toc_float: yes
always_allow_html: true
urlcolor: blue
mainfont: cochineal
sansfont: Fira Sans
monofont: Fira Code ## Although, see: https://tex.stackexchange.com/q/294362

## Automatically knit to both formats:
knit: (function(inputFile, encoding) {
 rmarkdown::render(inputFile, encoding = encoding, 
 output_format = 'all') 
 })
---

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position=c('top','right'),color='#33C1FF')
```


<style>
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    z-index: 2;
    color: #fff;
    background-color: #33C1FF;
    border-color: #97CAEF;
}

</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "",fig.align='center',message = FALSE,warning = FALSE)
require(here)
require(ggplot2)
require(plot3D)
require(kableExtra)
require(knitr)
require(giski)
require(magick)
options(scipen=99)

```

`r paste('[Updated:',format(Sys.time(),'%a, %b %d, %Y - %H:%M:%S'),']')`

<font color="black">

# 1. Overview

This tutorial demonstrates how to fit a finite mixture of piecewise mixed-effect models with unknown random knots using the R routine developed as described in [in our paper](https://github.com/czopluoglu/fitPMM/blob/main/paper/2015_Kohli%2C%20Harring%2C%20%26%20Zopluoglu.pdf). The R routine can be found in the [GitHub repository](https://github.com/czopluoglu/fitPMM) under the `/code` folder.

# 2. Installation and Setup

## 2.1 Download and Source the R Routine

First, download the R script file (fitPMM.R) from the repository and source it into your R environment.

```{r,eval=FALSE,echo=TRUE}
source('path_to_your_downloaded_code/fitPMM.R')

```

```{r,eval=TRUE,echo=FALSE}
source(here('code/fitPMM.R'))
```

## 2.2 Install and Load Required Libraries

The routine depends on several R packages. If these are not installed, use install.packages() to install them before loading.

```{r}
# Uncomment the lines to install if needed
# install.packages("segmented",dependencies=TRUE)
# install.packages("glmmML",dependencies=TRUE)
# install.packages("MCMCpack",dependencies=TRUE)
# install.packages("psych",dependencies=TRUE)
# install.packages("MBESS",dependencies=TRUE)
# install.packages("gdata",dependencies=TRUE)

# Load the necessary libraries
	require(segmented)   # Fitting piecewise model to individual data, used to obtain start values
	require(glmmML)      # Using "ghq()" function
	require(MCMCpack)    # Using "xpnd()" function        
	require(psych)       # Using "cor.smooth()" function
	require(MBESS)       # Using "cor2cov()" function
	require(gdata)       # Using "write.fwf()" function

```

# 3. Loading the dataset

We will use a random subsample of [The Early Childhood Longitudinal Study (ECLS-K)](http://nces.ed.gov/ecls/) . This dataset tracks children's progress from kindergarten through eighth grade. The time vector (`tvec`) represents specific intervals in the study.

First, download the data file from the folder [/data/](https://github.com/czopluoglu/fitPMM/tree/main/data) in the GitHub repository, and then load it into your R environment

```{r,eval=FALSE,echo=TRUE}
load(''path_to_your_downloaded_RData_file/ECLS.RData')  # Load the dataset
tvec <- c(0, 0.5, 1, 1.5, 3.5, 5.5, 8.5)  # Time points vector

head(ECLS)
```


```{r,eval=TRUE,echo=FALSE}
load(here('data/ECLS.RData'))  # Load the dataset
tvec <- c(0, 0.5, 1, 1.5, 3.5, 5.5, 8.5)  # Time points vector

head(ECLS)
```

The time points represent:

- 0   --- Kindergarten (fall)
- 0.5 --- Kindergarten (spring)
- 1   --- 1st Grade (fall)
- 1.5 --- 1st Grade (spring)
- 3.5 --- 3rd Grade (spring)
- 5.5 --- 5th Grade (spring)
- 8.5 --- 8th Grade (spring)

# 4. Generate starting values for the algorithm

The algorithm is sensitive to starting values. To improve convergence, we'll estimate initial parameters (intercepts, slopes, knots, and residual variance) by fitting piecewise models to each individual's data. These estimates will then be passed into the main model fitting function.

```{r}

# Time vector 
tvec = c(0,.5,1,1.5,3.5,5.5,8.5)  # Time vector

# Sample size, number of rows in the dataset

N <- nrow(ECLS) # Sample size						  

ind_par <- matrix(nrow=N,ncol=5)  # Matrix to store individual parameter estimates

colnames(ind_par) <- c("B1","B2","B3","K","Sigma")  

    # B1   : The intercept at the first phase
		# B2   : The slope at the first phase
		# B3   : The slope at the second phase
		# K    : Knot, change/break point between two phases
		# sigma: Residual Variance 
  
K.st <- 3		# Initial guess for the knot	  		

# The following loop may take some time depending on your sample size. 
# It fits piecewise linear model to each individual data, 
# and estimates B1, B2, B3, K, and sigma for each individual

for(i in 1:nrow(ECLS)) {    
  
  tryCatch({
		
    y <- as.numeric(ECLS[i,])
		
		if(sum(is.na(y))<=3){ # Ensure there are enough data points
		  
			lm.out <- lm(y~tvec)  
			
			out    <- segmented(lm.out,seg.Z=~tvec,psi=K.st) 
      
			if(is.null(out)!=TRUE & length(coef(out))==4) {
       			ind_par[i,1:2] <- as.numeric(coef(out))[1:2]    # B1, B2
	      		ind_par[i,3] <- ind_par[i,2]+as.numeric(coef(out))[3]  # B3
			      ind_par[i,4]<- out$psi[2]              # Knot
      			ind_par[i,5]<- (summary(out)$sigma)^2   # Residual variance
            	}
		}
  }, error=function(e){})
}

head(ind_par) 	   # Preview of individual estimates     
	
```


# 5. Fitting the model

Now, use the `fitPMM()` function to fit the mixture model. This function requires several arguments, including the dataset, starting values, and model-specific options such as the number of latent classes and quadrature points.

```{r}

fit <- fitPMM(
  data     = ECLS,     # Dataset
  start    = ind_par,  # Starting values
  time.vec = tvec,     # Time vector
  L        = 2,        # Number of latent classes
  nquad    = 20,       # Number of quadrature points
  isnit_max= 15,       # Max iterations for initial stage
  is_st    = 0.001,    # Stopping criteria for initial stage
  is_nset  = 500,      # Number of starting sets for initial stage
  fsnit_max= 2000,     # Max iterations for final stage
  fs_st    = 0.0001,   # Stopping criteria for final stage
  fs_nset  = 3,        # Number of starting sets for final stage
  sel      = 2,        # Optimization method (1 = BFGS, 2 = DFPQ)
  fix.cov  = TRUE,    # Fix variance-covariance across classes?
  fix.err  = TRUE     # Fix residual variance across classes?
)


```

# 6. Model output

The model will return several outputs, including parameter estimates and class memberships.

```{r}
fit$PAR         # Parameter estimates
```

```{r}
fit$CP          # Class proportions
```

```{r}
fit$POSTERIOR   # Posterior probabilities and class memberships
```

</font>




